---
title: "Winnipeg Assessment Parcels — 3D Value Map"
author: "Jason M. Schellenberg, P. App, AACI"
date: today
date-format: "MMMM D, YYYY"
format:
  html:
    toc: true
output-file: index.html
execute:
  echo: false
  warning: false
  message: false
---

## Overview

Interactive 3D map of Winnipeg's property assessment parcels.
**Height** encodes total assessed value; **color** encodes neighbourhood.
Hover over a parcel to see its assessed value, roll number, and how it compares to the neighbourhood median.
Use the **layer toggle** (bottom left) to switch to a neighbourhood-level view comparing total assessed value across neighbourhoods.

```{r}
#| label: setup
#| include: false

library(duckdb)
library(DBI)
library(freestiler)
library(mapgl)
library(arrow)
library(geoarrow)
library(dplyr)
library(sf)
library(ggplot2)
library(ggspatial)
library(scico)
library(gt)
library(fs)
library(glue)
library(cli)
library(scales)
library(readr)
library(stringr)

fs::dir_create("exports")

layer_name <- "parcels"
pmtiles_file <- "data/assessment-parcels.pmtiles"

# Set to TRUE to use local data embedded in the HTML (works offline, larger file).
# Set to FALSE to use PMTiles from docs/ (smaller file, requires Git LFS).
use_local_data <- FALSE

# PMTiles URL for the deployed map — relative path served from docs/ alongside index.html
pmtiles_url <- "assessment-parcels.pmtiles"

# Assessment roll metadata
base_value_date <- "April 1, 2023"
```

```{r}
#| label: find-data
#| include: false

parquet_files <- fs::dir_ls("data", glob = "*/assessment-parcels-*.parquet")

if (length(parquet_files) == 0) {
  stop("No GeoParquet files found in data/. Run WpgAssessmentParcels.qmd first.")
}

input_file <- sort(parquet_files, decreasing = TRUE)[1]
file_date <- stringr::str_extract(basename(input_file), "\\d{4}-\\d{2}-\\d{2}")

# Extract assessment year from the current_assessment_year field
asmt_year <- read_parquet(input_file, col_select = "current_assessment_year") |>
  pull(current_assessment_year) |>
  na.omit() |>
  head(1) |>
  as.character()

cli_alert_success("Using: {basename(input_file)}")
cli_alert_info("Download date: {file_date}")
cli_alert_info("File size: {fs::file_size(input_file)}")
```

## City Overview

```{r}
#| label: city-overview
#| include: false

parcels_all <- read_parquet(input_file) |> as.data.frame()

n_parcels <- nrow(parcels_all)
n_neighbourhoods <- n_distinct(parcels_all$neighbourhood_area)
total_assessed <- sum(parcels_all$total_assessed_value, na.rm = TRUE)
median_assessed <- median(parcels_all$total_assessed_value, na.rm = TRUE)

# Top neighbourhood by total assessed value
top_neighbourhood <- parcels_all |>
  group_by(neighbourhood_area) |>
  summarise(total = sum(total_assessed_value, na.rm = TRUE), .groups = "drop") |>
  slice_max(total, n = 1)
```

Winnipeg's assessment roll (`r asmt_year` — `r base_value_date` reference date) covers **`r format(n_parcels, big.mark = ",")`** property parcels across **`r n_neighbourhoods`** neighbourhoods, with a combined assessed value of **`r dollar(total_assessed, accuracy = 1, scale = 1e-9, suffix = "B")`**. The median property assessment is **`r dollar(median_assessed, accuracy = 1)`**. The largest neighbourhood by total assessed value is **`r top_neighbourhood$neighbourhood_area`** at `r dollar(top_neighbourhood$total, accuracy = 1, scale = 1e-9, suffix = "B")`.

Data source: [City of Winnipeg Assessment Parcels](https://data.winnipeg.ca/Assessment-Taxation-Corporate/Assessment-Parcels/d4mq-wa44) (downloaded `r file_date`).

```{r}
#| label: load-map-data
#| include: false

cli_alert_info("Loading parcel geometry from {basename(input_file)}...")

parcels_sf <- read_parquet(input_file) |>
  mutate(geometry = st_as_sfc(geometry)) |>
  st_as_sf()

# Simplify geometry for faster rendering and smaller HTML output
# dTolerance = 0.0002 (~22m) keeps parcels recognizable at street zoom
if (use_local_data) {
  sf_use_s2(FALSE)
  parcels_sf <- st_make_valid(parcels_sf) |>
    st_simplify(dTolerance = 0.0002, preserveTopology = TRUE)
  sf_use_s2(TRUE)
  cli_alert_info("Simplified geometry for local embedding")
}

# Compute neighbourhood medians and join back for tooltip context
neighbourhood_medians <- parcels_sf |>
  st_drop_geometry() |>
  group_by(neighbourhood_area) |>
  summarise(Neighbourhood_Median = median(total_assessed_value, na.rm = TRUE), .groups = "drop")

parcels_sf <- parcels_sf |>
  left_join(neighbourhood_medians, by = "neighbourhood_area") |>
  mutate(
    Neighbourhood_Median_Fmt = if_else(
      is.na(Neighbourhood_Median), "N/A", dollar(Neighbourhood_Median, accuracy = 1)
    ),
    Total_Value_Fmt = dollar(total_assessed_value, accuracy = 1),
    # Compute per-parcel % change for the 3D change map
    # Exclude outliers: proposed = $0 (demolitions/exemptions) and current = $1 (placeholder values)
    pct_change = if_else(
      !is.na(total_assessed_value) & !is.na(total_proposed_assessment_value) &
        total_assessed_value > 1 & total_proposed_assessment_value > 0,
      (total_proposed_assessment_value - total_assessed_value) / total_assessed_value * 100,
      NA_real_
    ),
    pct_change_fmt = if_else(is.na(pct_change), "N/A", paste0(round(pct_change, 1), "%")),
    # Clamped version for 3D extrusion height (cap at 100% to avoid extreme spikes)
    pct_change_clamped = pmin(pmax(pct_change, 0), 100),
    # Broad property use category from the prefix of the use code
    use_category = case_when(
      str_starts(property_use_code, "RES") ~ "Residential",
      str_starts(property_use_code, "CN")  ~ "Condo",
      str_starts(property_use_code, "CM")  ~ "Commercial",
      str_starts(property_use_code, "IN")  ~ "Industrial",
      str_starts(property_use_code, "PII") ~ "Institutional",
      str_starts(property_use_code, "PIR") ~ "Recreational",
      str_starts(property_use_code, "AGR") ~ "Agricultural",
      str_starts(property_use_code, "V")   ~ "Vacant",
      TRUE                                 ~ "Other"
    )
  ) |>
  select(roll_number, total_assessed_value, Total_Value_Fmt,
         total_proposed_assessment_value, pct_change, pct_change_clamped, pct_change_fmt,
         Neighbourhood_Median, Neighbourhood_Median_Fmt,
         full_address, neighbourhood_area,
         property_use_code, use_category, building_type, year_built,
         total_living_area, dwelling_units)

cli_alert_success("Loaded {format(nrow(parcels_sf), big.mark = ',')} parcels")

# Dissolve parcels into neighbourhood boundaries for choropleth layer
cli_alert_info("Dissolving parcels by neighbourhood...")
sf_use_s2(FALSE)
parcels_sf <- st_make_valid(parcels_sf)
neighbourhood_sf <- parcels_sf |>
  group_by(neighbourhood_area) |>
  summarise(
    Total_Assessed = sum(total_assessed_value, na.rm = TRUE),
    Parcel_Count = n(),
    Median_Value = median(total_assessed_value, na.rm = TRUE),
    .groups = "drop"
  ) |>
  st_make_valid() |>
  st_simplify(dTolerance = 0.001, preserveTopology = TRUE) |>
  st_make_valid() |>
  mutate(
    Total_Assessed_Fmt = dollar(Total_Assessed, accuracy = 1),
    Median_Value_Fmt = dollar(Median_Value, accuracy = 1),
    Value_Rank = paste(rank(-Total_Assessed, ties.method = "min"), "of", n())
  )
sf_use_s2(TRUE)
cli_alert_success("Created {nrow(neighbourhood_sf)} neighbourhood polygons")
```

```{r}
#| label: generate-pmtiles
#| include: false
#| eval: !expr '!use_local_data'

# Skip if PMTiles is newer than source parquet
needs_rebuild <- !fs::file_exists(pmtiles_file) ||
  fs::file_info(pmtiles_file)$modification_time <
  fs::file_info(input_file)$modification_time

if (needs_rebuild) {
  cli_alert_info("Generating PMTiles from {basename(input_file)}...")

  freestile(
    input = parcels_sf,
    output = pmtiles_file,
    layer_name = layer_name,
    min_zoom = 8L,
    max_zoom = 14L
  )

  cli_alert_success("PMTiles written: {fs::file_size(pmtiles_file)}")
} else {
  cli_alert_info("PMTiles up to date: {fs::file_size(pmtiles_file)}")
}
```

```{r}
#| label: neighbourhood-codes
#| include: false

parcels_df <- parcels_all |>
  count(neighbourhood_area, sort = TRUE)

cli_alert_info("Found {nrow(parcels_df)} neighbourhoods")
```

```{r}
#| label: define-palette
#| include: false

# Define a categorical color palette for the top neighbourhoods
# Colors chosen for visual distinction on a dark basemap
palette <- c(
  "#e74c3c",   # red
  "#3498db",   # blue
  "#2ecc71",   # green
  "#f39c12",   # orange
  "#9b59b6",   # purple
  "#1abc9c",   # teal
  "#e67e22",   # dark orange
  "#ec407a",   # pink
  "#26c6da",   # cyan
  "#ffca28",   # amber
  "#8d6e63",   # brown
  "#78909c"    # blue-grey
)

# Use top neighbourhoods by parcel count for color coding
neighbourhood_names <- parcels_df$neighbourhood_area
neighbourhood_colors <- rep_len(palette, length(neighbourhood_names))

# Property use broad categories and colors for the change map
use_category_map <- c(
  "RES" = "Residential",
  "CN"  = "Condo",
  "CM"  = "Commercial",
  "IN"  = "Industrial",
  "PII" = "Institutional",
  "PIR" = "Recreational",
  "AGR" = "Agricultural",
  "V"   = "Vacant"
)

use_category_colors <- c(
  "Residential"  = "#e74c3c",
  "Condo"        = "#3498db",
  "Commercial"   = "#f39c12",
  "Industrial"   = "#9b59b6",
  "Institutional" = "#1abc9c",
  "Recreational"  = "#2ecc71",
  "Agricultural"  = "#8d6e63",
  "Vacant"       = "#78909c",
  "Other"        = "#95a5a6"
)
```

## Assessed Value by Parcel

Each parcel is extruded in 3D — **taller buildings represent higher assessed values**. Parcels are coloured by neighbourhood. Hover over any parcel to see its address, assessed value, and how it compares to the neighbourhood median. Use the **layer toggle** (bottom left) to switch between the parcel-level 3D view and a neighbourhood-level choropleth of total assessed value.

```{r}
#| label: map-3d

# Quantile breaks for neighbourhood choropleth color scale
total_vals <- neighbourhood_sf$Total_Assessed[!is.na(neighbourhood_sf$Total_Assessed)]
q_breaks <- as.numeric(quantile(total_vals, probs = c(0, 0.2, 0.4, 0.6, 0.8, 1.0)))

m <- maplibre(
  style = carto_style("positron"),
  center = c(-97.14, 49.89),
  zoom = 11,
  pitch = 30,
  bearing = 0,
  projection = "mercator"
) |>
  add_navigation_control(position = "top-right")

# Neighbourhood choropleth layer (hidden by default)
m <- m |>
  add_fill_layer(
    id = "neighbourhood-total-value",
    source = neighbourhood_sf,
    fill_color = interpolate(
      column = "Total_Assessed",
      values = q_breaks,
      stops = c("#2166ac", "#67a9cf", "#d1e5f0", "#fddbc7", "#ef8a62", "#b2182b")
    ),
    fill_opacity = 0.8,
    visibility = "none",
    tooltip = concat(
      "<strong>", get_column("neighbourhood_area"), "</strong><br>",
      "Total Assessed Value: ", get_column("Total_Assessed_Fmt"), "<br>",
      "Rank: ", get_column("Value_Rank"), "<br>",
      "Parcels: ", get_column("Parcel_Count"), "<br>",
      "Median Value: ", get_column("Median_Value_Fmt")
    ),
    hover_options = list(
      fill_color = "#ffffff",
      fill_opacity = 1
    )
  )

# 3D parcel extrusion layer (visible by default)
if (use_local_data) {
  m <- m |>
    add_fill_extrusion_layer(
      id = "wpg-parcels-3d",
      source = parcels_sf,
      fill_extrusion_height = interpolate(
        column = "total_assessed_value",
        values = c(0, 100000, 500000, 2000000, 10000000),
        stops = c(0, 50, 200, 600, 2000)
      ),
      fill_extrusion_color = match_expr(
        column = "neighbourhood_area",
        values = neighbourhood_names,
        stops = neighbourhood_colors,
        default = "#95a5a6"
      ),
      fill_extrusion_opacity = 0.8,
      tooltip = concat(
        "<strong>", get_column("full_address"), "</strong><br>",
        "Assessed Value: ", get_column("Total_Value_Fmt"), "<br>",
        "Neighbourhood Median: ", get_column("Neighbourhood_Median_Fmt"), "<br>",
        "Neighbourhood: ", get_column("neighbourhood_area"), "<br>",
        "Roll No: ", get_column("roll_number"), "<br>",
        "Use: ", get_column("property_use_code"), "<br>",
        "Dwelling Units: ", get_column("dwelling_units")
      ),
      hover_options = list(
        fill_extrusion_color = "#ffffff"
      )
    )
} else {
  m <- m |>
    add_pmtiles_source(
      id = "wpg-parcels-source",
      url = pmtiles_url
    ) |>
    add_fill_extrusion_layer(
      id = "wpg-parcels-3d",
      source = "wpg-parcels-source",
      source_layer = layer_name,
      fill_extrusion_height = interpolate(
        column = "total_assessed_value",
        values = c(0, 100000, 500000, 2000000, 10000000),
        stops = c(0, 50, 200, 600, 2000)
      ),
      fill_extrusion_color = match_expr(
        column = "neighbourhood_area",
        values = neighbourhood_names,
        stops = neighbourhood_colors,
        default = "#95a5a6"
      ),
      fill_extrusion_opacity = 0.8,
      tooltip = concat(
        "<strong>", get_column("full_address"), "</strong><br>",
        "Assessed Value: ", get_column("Total_Value_Fmt"), "<br>",
        "Neighbourhood Median: ", get_column("Neighbourhood_Median_Fmt"), "<br>",
        "Neighbourhood: ", get_column("neighbourhood_area"), "<br>",
        "Roll No: ", get_column("roll_number"), "<br>",
        "Use: ", get_column("property_use_code"), "<br>",
        "Dwelling Units: ", get_column("dwelling_units")
      ),
      hover_options = list(
        fill_extrusion_color = "#ffffff"
      )
    )
}

# Build legend with abbreviated breakpoint labels (e.g. $1.2B, $450M)
legend_colors <- c("#2166ac", "#67a9cf", "#d1e5f0", "#fddbc7", "#ef8a62", "#b2182b")
fmt_short <- function(x) {
  dplyr::case_when(
    x >= 1e9  ~ paste0("$", trimws(format(round(x / 1e9, 1), nsmall = 1)), "B"),
    x >= 1e6  ~ paste0("$", trimws(format(round(x / 1e6, 0))), "M"),
    x >= 1e3  ~ paste0("$", trimws(format(round(x / 1e3, 0))), "K"),
    TRUE      ~ paste0("$", trimws(format(round(x, 0))))
  )
}
legend_labels <- fmt_short(q_breaks)

legend_rows <- paste0(sapply(seq_along(legend_colors), function(i) {
  label <- if (i == 1) {
    paste0("&lt; ", legend_labels[i + 1])
  } else if (i == length(legend_colors)) {
    paste0("&ge; ", legend_labels[i])
  } else {
    paste0(legend_labels[i], " \u2013 ", legend_labels[i + 1])
  }
  paste0(
    '<div style="display:flex;align-items:center;margin-bottom:3px;">',
    '<span style="display:inline-block;width:18px;height:12px;background:', legend_colors[i],
    ';border-radius:2px;margin-right:6px;flex-shrink:0;"></span>',
    '<span>', label, '</span></div>'
  )
}), collapse = "")

# Combine layer toggle + legend into ONE add_control (mapgl only keeps one custom control)
combined_control_html <- paste0(
  '<div style="background:rgba(255,255,255,0.85);padding:10px 12px;border-radius:6px;color:#333;font-family:sans-serif;font-size:13px;line-height:1.6;border:1px solid #ccc;">',
  '<div style="font-weight:bold;margin-bottom:4px;">View</div>',
  '<label style="display:block;cursor:pointer;">',
  '<input type="radio" name="map-layer" value="3d" checked style="margin-right:5px;">3D Parcels',
  '</label>',
  '<label style="display:block;cursor:pointer;">',
  '<input type="radio" name="map-layer" value="neighbourhood" style="margin-right:5px;">Total Value by Neighbourhood',
  '</label>',
  '<div id="neighbourhood-legend" style="display:none;margin-top:8px;padding-top:8px;border-top:1px solid rgba(0,0,0,0.2);font-size:12px;line-height:1.5;">',
  '<div style="font-weight:bold;margin-bottom:5px;">Total Assessed Value</div>',
  legend_rows,
  '</div>',
  '</div>'
)

m <- m |>
  add_control(combined_control_html, position = "bottom-left") |>
  htmlwidgets::onRender("
    function(el, x) {
      var poll = setInterval(function() {
        if (!el.map || !el.map.isStyleLoaded()) return;
        clearInterval(poll);
        var map = el.map;
        var legend = document.getElementById('neighbourhood-legend');
        var radios = el.querySelectorAll('input[name=\"map-layer\"]');
        radios.forEach(function(radio) {
          radio.addEventListener('change', function() {
            if (this.value === '3d') {
              map.setLayoutProperty('wpg-parcels-3d', 'visibility', 'visible');
              map.setLayoutProperty('neighbourhood-total-value', 'visibility', 'none');
              if (legend) legend.style.display = 'none';
            } else {
              map.setLayoutProperty('wpg-parcels-3d', 'visibility', 'none');
              map.setLayoutProperty('neighbourhood-total-value', 'visibility', 'visible');
              if (legend) legend.style.display = 'block';
            }
          });
        });
      }, 200);
    }
  ")

m
```

## Assessed Value by Neighbourhood (`r asmt_year`)

Click any column header to sort.

```{r}
#| label: summary-stats

library(reactable)

summary_table <- parcels_all |>
  group_by(neighbourhood_area) |>
  summarise(
    count = n(),
    total_value = sum(total_assessed_value, na.rm = TRUE),
    median_value = median(total_assessed_value, na.rm = TRUE),
    mean_value = mean(total_assessed_value, na.rm = TRUE),
    max_value = if (all(is.na(total_assessed_value))) NA_real_
                else max(total_assessed_value, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(total_value))

dollar_col <- function(header) {
  colDef(
    name = header,
    format = colFormat(prefix = "$", separators = TRUE, digits = 0),
    defaultSortOrder = "desc"
  )
}

reactable(
  summary_table,
  sortable = TRUE,
  defaultSorted = list(total_value = "desc"),
  defaultPageSize = 25,
  showPageSizeOptions = TRUE,
  pageSizeOptions = c(25, 50, 100),
  striped = TRUE,
  highlight = TRUE,
  compact = TRUE,
  searchable = TRUE,
  columns = list(
    neighbourhood_area = colDef(name = "Neighbourhood", minWidth = 170),
    count = colDef(
      name = "Parcels",
      format = colFormat(separators = TRUE),
      defaultSortOrder = "desc"
    ),
    total_value  = colDef(
      name = "Total Assessment",
      format = colFormat(prefix = "$", separators = TRUE, digits = 0),
      defaultSortOrder = "desc",
      minWidth = 145
    ),
    median_value = dollar_col("Median"),
    mean_value   = dollar_col("Mean"),
    max_value    = colDef(name = "Max", format = colFormat(prefix = "$", separators = TRUE, digits = 0), defaultSortOrder = "desc", minWidth = 110)
  )
)
```

```{r}
#| label: compute-change
#| include: false

# Compute per-parcel % change from current (2026) to proposed (2027)
change_stats <- parcels_all |>
  filter(
    !is.na(total_assessed_value),
    !is.na(total_proposed_assessment_value),
    total_assessed_value > 0
  ) |>
  mutate(
    pct_change = (total_proposed_assessment_value - total_assessed_value) /
                  total_assessed_value * 100
  )

n_change_parcels <- nrow(change_stats)
citywide_median_change <- median(change_stats$pct_change, na.rm = TRUE)

# Summarize by neighbourhood
change_by_neighbourhood <- change_stats |>
  group_by(neighbourhood_area) |>
  summarise(
    change_parcel_count = n(),
    median_pct          = median(pct_change, na.rm = TRUE),
    mean_pct            = mean(pct_change, na.rm = TRUE),
    iqr_pct             = IQR(pct_change, na.rm = TRUE),
    q25_pct             = quantile(pct_change, 0.25, na.rm = TRUE),
    q75_pct             = quantile(pct_change, 0.75, na.rm = TRUE),
    .groups = "drop"
  )

# Join change stats to existing dissolved neighbourhood boundaries
change_neighbourhood_sf <- neighbourhood_sf |>
  left_join(change_by_neighbourhood, by = "neighbourhood_area") |>
  mutate(
    median_pct_fmt = paste0(round(median_pct, 1), "%"),
    mean_pct_fmt   = paste0(round(mean_pct, 1), "%"),
    iqr_pct_fmt    = paste0(round(iqr_pct, 1), "%")
  )

cli_alert_info("Citywide median change: {round(citywide_median_change, 1)}%")
```

## Assessment Change — 2026 → 2027 (Proposed)

This section compares Winnipeg's **current (`r asmt_year`)** property assessments to the **proposed (2027)** values across `r format(n_change_parcels, big.mark = ",")` parcels. The citywide median assessment change is **`r round(citywide_median_change, 1)`%**.

**Note on data quality:** The dataset contains approximately 8,800 parcels where the proposed 2027 assessment is \$0, producing a -100% change. These are likely demolitions, tax exemptions, or parcels being re-rolled rather than genuine value declines. Similarly, a small number of parcels have a nominal current assessment of \$1 (placeholder values) with substantial proposed values, producing extreme percentage increases. Both categories are excluded from the parcel-level 3D change map below but remain in the neighbourhood-level median calculations (where they have minimal impact due to the use of medians).

### Where Assessments Jumped Most

Median percentage change in assessed value by neighbourhood. The diverging scale is centred on the citywide median (`r round(citywide_median_change, 1)`%). Neighbourhoods that show a median change of -100% are likely based on incomplete data — typically areas with very few parcels where most have a proposed assessment of $0.

#### Interactive Map

```{r}
#| label: map-a-mapgl

# Diverging breaks centred on citywide median
median_vals <- change_neighbourhood_sf$median_pct[!is.na(change_neighbourhood_sf$median_pct)]
val_range <- range(median_vals)
max_dev <- max(abs(val_range - citywide_median_change))
break_low  <- citywide_median_change - max_dev
break_high <- citywide_median_change + max_dev
q_breaks_a <- c(break_low,
                citywide_median_change - max_dev * 0.5,
                citywide_median_change,
                citywide_median_change + max_dev * 0.5,
                break_high)

maplibre(
  style = carto_style("positron"),
  center = c(-97.14, 49.89),
  zoom = 11,
  pitch = 0,
  bearing = 0,
  projection = "mercator"
) |>
  add_navigation_control(position = "top-right") |>
  add_fill_layer(
    id = "median-change",
    source = change_neighbourhood_sf,
    fill_color = interpolate(
      column = "median_pct",
      values = q_breaks_a,
      stops = c("#2166ac", "#92c5de", "#f7f7f7", "#f4a582", "#b2182b")
    ),
    fill_opacity = 0.85,
    tooltip = concat(
      "<strong>", get_column("neighbourhood_area"), "</strong><br>",
      "Median Change: ", get_column("median_pct_fmt"), "<br>",
      "Mean Change: ", get_column("mean_pct_fmt"), "<br>",
      "Parcels: ", get_column("change_parcel_count")
    ),
    hover_options = list(
      fill_color = "#ffffff",
      fill_opacity = 1
    )
  ) |>
  add_line_layer(
    id = "neighbourhood-borders-a",
    source = change_neighbourhood_sf,
    line_color = "#333333",
    line_width = 0.5,
    line_opacity = 0.4
  )
```

### Assessment Change by Parcel

Each parcel is extruded in 3D — **taller parcels had larger percentage increases** in assessed value from `r asmt_year` to the proposed 2027 roll. **Colour encodes property use category** (residential, condo, commercial, industrial, institutional, recreational, agricultural, vacant). Hover to see the address, current value, and percentage change.

**Outlier filtering applied:** Approximately 8,800 parcels with a proposed 2027 assessment of \$0 are excluded — these represent demolitions, tax exemptions, or parcels being re-rolled and would otherwise all show as -100% change. A small number of parcels with a nominal current assessment of \$1 (placeholder values) are also excluded, as they produce misleadingly extreme percentage increases. For the remaining parcels, extrusion heights are capped at 100% change to keep the visualization readable; the actual percentage is still shown in the tooltip on hover.

```{r}
#| label: map-change-3d

use_cat_names  <- names(use_category_colors)
use_cat_colors <- unname(use_category_colors)

if (use_local_data) {
  m_change <- maplibre(
    style = carto_style("positron"),
    center = c(-97.14, 49.89),
    zoom = 11,
    pitch = 45,
    bearing = -15,
    projection = "mercator"
  ) |>
    add_navigation_control(position = "top-right") |>
    add_fill_extrusion_layer(
      id = "wpg-change-3d",
      source = parcels_sf,
      fill_extrusion_height = interpolate(
        column = "pct_change_clamped",
        values = c(0, 10, 25, 50, 75, 100),
        stops  = c(0, 80, 250, 500, 750, 1000)
      ),
      fill_extrusion_color = match_expr(
        column = "use_category",
        values = use_cat_names,
        stops = use_cat_colors,
        default = "#95a5a6"
      ),
      fill_extrusion_opacity = 0.8,
      tooltip = concat(
        "<strong>", get_column("full_address"), "</strong><br>",
        "Change: ", get_column("pct_change_fmt"), "<br>",
        "Current (2026): ", get_column("Total_Value_Fmt"), "<br>",
        "Use: ", get_column("property_use_code"), "<br>",
        "Neighbourhood: ", get_column("neighbourhood_area"), "<br>",
        "Roll No: ", get_column("roll_number")
      ),
      hover_options = list(
        fill_extrusion_color = "#ffffff"
      )
    )
} else {
  m_change <- maplibre(
    style = carto_style("positron"),
    center = c(-97.14, 49.89),
    zoom = 11,
    pitch = 45,
    bearing = -15,
    projection = "mercator"
  ) |>
    add_navigation_control(position = "top-right") |>
    add_pmtiles_source(
      id = "wpg-change-source",
      url = pmtiles_url
    ) |>
    add_fill_extrusion_layer(
      id = "wpg-change-3d",
      source = "wpg-change-source",
      source_layer = layer_name,
      fill_extrusion_height = interpolate(
        column = "pct_change_clamped",
        values = c(0, 10, 25, 50, 75, 100),
        stops  = c(0, 80, 250, 500, 750, 1000)
      ),
      fill_extrusion_color = match_expr(
        column = "use_category",
        values = use_cat_names,
        stops = use_cat_colors,
        default = "#95a5a6"
      ),
      fill_extrusion_opacity = 0.8,
      tooltip = concat(
        "<strong>", get_column("full_address"), "</strong><br>",
        "Change: ", get_column("pct_change_fmt"), "<br>",
        "Current (2026): ", get_column("Total_Value_Fmt"), "<br>",
        "Use: ", get_column("property_use_code"), "<br>",
        "Neighbourhood: ", get_column("neighbourhood_area"), "<br>",
        "Roll No: ", get_column("roll_number")
      ),
      hover_options = list(
        fill_extrusion_color = "#ffffff"
      )
    )
}

# Add legend for use categories
use_legend_rows <- paste0(sapply(seq_along(use_cat_names), function(i) {
  paste0(
    '<div style="display:flex;align-items:center;margin-bottom:3px;">',
    '<span style="display:inline-block;width:18px;height:12px;background:', use_cat_colors[i],
    ';border-radius:2px;margin-right:6px;flex-shrink:0;"></span>',
    '<span>', use_cat_names[i], '</span></div>'
  )
}), collapse = "")

use_legend_html <- paste0(
  '<div style="background:rgba(255,255,255,0.85);padding:10px 12px;border-radius:6px;color:#333;font-family:sans-serif;font-size:12px;line-height:1.5;border:1px solid #ccc;">',
  '<div style="font-weight:bold;margin-bottom:5px;">Property Use</div>',
  use_legend_rows,
  '</div>'
)

m_change <- m_change |>
  add_control(use_legend_html, position = "bottom-left")

m_change
```

### Total Assessment and Change by Property Use

The total assessed value of all properties in Winnipeg for `r asmt_year` is **`r scales::dollar(total_assessed, accuracy = 1)`**. Median percentage change excludes outliers. Q25 and Q75 are the 25th and 75th percentiles. Click any column header to sort. This highlights how different property types are affected by the reassessment.

```{r}
#| label: change-by-use

library(reactable)

# Filter to same outlier-excluded set used in the 3D map
change_by_use <- parcels_all |>
  filter(
    !is.na(total_assessed_value),
    !is.na(total_proposed_assessment_value),
    total_assessed_value > 1,
    total_proposed_assessment_value > 0
  ) |>
  mutate(
    pct_change = (total_proposed_assessment_value - total_assessed_value) /
                  total_assessed_value * 100
  ) |>
  group_by(property_use_code) |>
  summarise(
    parcel_count    = n(),
    total_assessed  = sum(total_assessed_value, na.rm = TRUE),
    median_pct      = round(median(pct_change, na.rm = TRUE), 1),
    mean_pct        = round(mean(pct_change, na.rm = TRUE), 1),
    q25_pct         = round(quantile(pct_change, 0.25, na.rm = TRUE), 1),
    q75_pct         = round(quantile(pct_change, 0.75, na.rm = TRUE), 1),
    .groups = "drop"
  ) |>
  slice_max(parcel_count, n = 50) |>
  arrange(desc(median_pct))

# Color scale for median column
median_change_pal <- scales::col_numeric(
  palette = c("#2166ac", "#f7f7f7", "#b2182b"),
  domain = range(change_by_use$median_pct, na.rm = TRUE)
)

pct_col_def <- function(header) {
  colDef(
    name = header,
    format = colFormat(suffix = "%", digits = 1),
    defaultSortOrder = "desc"
  )
}

reactable(
  change_by_use,
  sortable = TRUE,
  defaultSorted = list(total_assessed = "desc"),
  defaultPageSize = 25,
  showPageSizeOptions = TRUE,
  pageSizeOptions = c(25, 50),
  striped = TRUE,
  highlight = TRUE,
  compact = TRUE,
  columns = list(
    property_use_code = colDef(name = "Property Use", minWidth = 200),
    parcel_count = colDef(
      name = "Parcels",
      format = colFormat(separators = TRUE),
      defaultSortOrder = "desc"
    ),
    total_assessed = colDef(
      name = "Total Assessed",
      format = colFormat(prefix = "$", separators = TRUE, digits = 0),
      defaultSortOrder = "desc",
      minWidth = 155
    ),
    median_pct = colDef(
      name = "Median %",
      format = colFormat(suffix = "%", digits = 1),
      defaultSortOrder = "desc",
      minWidth = 95,
      style = function(value) {
        list(background = median_change_pal(value), fontWeight = "bold")
      }
    ),
    mean_pct = colDef(show = FALSE),
    q25_pct  = colDef(name = "Q25", format = colFormat(suffix = "%", digits = 1), defaultSortOrder = "desc", maxWidth = 75),
    q75_pct  = colDef(name = "Q75", format = colFormat(suffix = "%", digits = 1), defaultSortOrder = "desc", maxWidth = 75)
  )
)
```

## Residential Property Characteristics

This section focuses on residential properties (RESSD, RESSS, RESDU, RESMA, RESRH, RESSU, CNAPT, CNRES) where both year built and living area data are available. Use the **Metric** toggle to switch between Year Built and Living Area, and the **View** toggle to switch between individual 3D parcels and neighbourhood-level medians. In 3D mode, **height encodes the selected metric** while **colour encodes the other metric** — for example, when viewing Year Built, taller parcels are newer and colour shows living area.

```{r}
#| label: res-data
#| include: false

res_code_prefixes <- c("RESSD", "CNAPT", "RESSS", "CNRES", "RESDU", "RESMA", "RESRH", "RESSU")

# Property use codes include descriptions (e.g. "RESSD - DETACHED SINGLE DWELLING")
# Build a regex pattern to match any of the prefixes at the start
res_pattern <- paste0("^(", paste(res_code_prefixes, collapse = "|"), ")")

res_sf <- parcels_sf |>
  filter(
    str_detect(property_use_code, res_pattern),
    !is.na(year_built), year_built > 0,
    !is.na(total_living_area), total_living_area > 0
  ) |>
  mutate(
    res_use_code = str_extract(property_use_code, "^[A-Z]+")
  )

cli_alert_info("Residential parcels with year built & living area: {nrow(res_sf)}")

# Get the actual full property_use_code values for PMTiles filter
res_full_codes <- unique(res_sf$property_use_code)

res_colours <- c(
  "RESSD" = "#e74c3c",
  "RESSS" = "#2ecc71",
  "RESDU" = "#9b59b6",
  "RESMA" = "#e67e22",
  "RESRH" = "#1abc9c",
  "RESSU" = "#34495e",
  "CNAPT" = "#3498db",
  "CNRES" = "#f1c40f"
)

# Dissolve to neighbourhood polygons with residential stats
sf_use_s2(FALSE)
res_sf <- st_make_valid(res_sf)
res_neighbourhood_sf <- res_sf |>
  group_by(neighbourhood_area) |>
  summarise(
    Parcel_Count       = n(),
    Median_Year_Built  = median(year_built, na.rm = TRUE),
    Mean_Year_Built    = round(mean(year_built, na.rm = TRUE)),
    Median_Living_Area = median(total_living_area, na.rm = TRUE),
    Mean_Living_Area   = round(mean(total_living_area, na.rm = TRUE)),
    Total_Assessed     = sum(total_assessed_value, na.rm = TRUE),
    .groups = "drop"
  ) |>
  st_make_valid() |>
  st_simplify(dTolerance = 0.001, preserveTopology = TRUE) |>
  st_make_valid() |>
  mutate(
    Total_Assessed_Fmt     = dollar(Total_Assessed, accuracy = 1),
    Median_Living_Area_Fmt = format(round(Median_Living_Area), big.mark = ","),
    Mean_Living_Area_Fmt   = format(round(Mean_Living_Area), big.mark = ",")
  )
sf_use_s2(TRUE)

cli_alert_success("Residential neighbourhood dissolve: {nrow(res_neighbourhood_sf)} neighbourhoods")

# Quantile breaks for choropleths
yb_vals <- res_neighbourhood_sf$Median_Year_Built[!is.na(res_neighbourhood_sf$Median_Year_Built)]
yb_breaks <- as.numeric(quantile(yb_vals, probs = c(0, 0.2, 0.4, 0.6, 0.8, 1.0)))
yb_colors <- c("#ffffcc", "#c7e9b4", "#7fcdbb", "#41b6c4", "#1d91c0", "#225ea8")

la_vals <- res_neighbourhood_sf$Median_Living_Area[!is.na(res_neighbourhood_sf$Median_Living_Area)]
la_breaks <- as.numeric(quantile(la_vals, probs = c(0, 0.2, 0.4, 0.6, 0.8, 1.0)))
la_colors <- c("#feebe2", "#fcc5c0", "#fa9fb5", "#f768a1", "#c51b8a", "#7a0177")
```

```{r}
#| label: res-map

# --- Parcel-level colour scales for year built and living area ---
# Year built: older = cool/light, newer = warm/dark
yb_parcel_colors <- c("#ffffcc", "#c7e9b4", "#7fcdbb", "#41b6c4", "#1d91c0", "#225ea8")
yb_parcel_breaks <- c(1900, 1950, 1970, 1985, 2000, 2015)

# Living area: smaller = light, larger = dark
la_parcel_colors <- c("#feebe2", "#fcc5c0", "#fa9fb5", "#f768a1", "#c51b8a", "#7a0177")
la_parcel_breaks <- c(500, 800, 1100, 1500, 2200, 4000)

# Shared tooltip for all 3D layers
res_tooltip <- concat(
  "<strong>", get_column("full_address"), "</strong><br>",
  "Year Built: ", get_column("year_built"), "<br>",
  "Living Area: ", get_column("total_living_area"), " sq ft<br>",
  "Use: ", get_column("property_use_code"), "<br>",
  "Assessed Value: ", get_column("Total_Value_Fmt"), "<br>",
  "Neighbourhood: ", get_column("neighbourhood_area")
)

# PMTiles filter for residential codes
res_pmtiles_filter <- list("all",
  list("in", list("get", "property_use_code"), list("literal", as.list(res_full_codes))),
  list(">", list("get", "year_built"), 0),
  list(">", list("get", "total_living_area"), 0)
)

# --- Map base ---
m_res <- maplibre(
  style = carto_style("positron"),
  center = c(-97.14, 49.89),
  zoom = 11,
  pitch = 30,
  bearing = 0,
  projection = "mercator"
) |>
  add_navigation_control(position = "top-right")

# --- Neighbourhood choropleth layers (both hidden by default) ---

# Year Built choropleth
m_res <- m_res |>
  add_fill_layer(
    id = "res-nb-yearbuilt",
    source = res_neighbourhood_sf,
    fill_color = interpolate(
      column = "Median_Year_Built",
      values = yb_breaks,
      stops = yb_colors
    ),
    fill_opacity = 0.8,
    visibility = "none",
    tooltip = concat(
      "<strong>", get_column("neighbourhood_area"), "</strong><br>",
      "Median Year Built: ", get_column("Median_Year_Built"), "<br>",
      "Mean Year Built: ", get_column("Mean_Year_Built"), "<br>",
      "Residential Parcels: ", get_column("Parcel_Count")
    ),
    hover_options = list(fill_color = "#ffffff", fill_opacity = 1)
  )

# Living Area choropleth
m_res <- m_res |>
  add_fill_layer(
    id = "res-nb-livingarea",
    source = res_neighbourhood_sf,
    fill_color = interpolate(
      column = "Median_Living_Area",
      values = la_breaks,
      stops = la_colors
    ),
    fill_opacity = 0.8,
    visibility = "none",
    tooltip = concat(
      "<strong>", get_column("neighbourhood_area"), "</strong><br>",
      "Median Living Area: ", get_column("Median_Living_Area_Fmt"), " sq ft<br>",
      "Mean Living Area: ", get_column("Mean_Living_Area_Fmt"), " sq ft<br>",
      "Residential Parcels: ", get_column("Parcel_Count")
    ),
    hover_options = list(fill_color = "#ffffff", fill_opacity = 1)
  )

# --- 3D Parcel layers (4 combinations, only one visible at a time) ---
# Default: height=year_built, colour=living_area

if (use_local_data) {
  # 3D: height=year_built, colour=living_area (DEFAULT - visible)
  m_res <- m_res |>
    add_fill_extrusion_layer(
      id = "res-3d-yb",
      source = res_sf,
      fill_extrusion_height = interpolate(
        column = "year_built",
        values = c(1900, 1940, 1960, 1980, 2000, 2025),
        stops  = c(20, 80, 200, 400, 700, 1000)
      ),
      fill_extrusion_color = interpolate(
        column = "total_living_area",
        values = la_parcel_breaks,
        stops  = la_parcel_colors
      ),
      fill_extrusion_opacity = 0.8,
      tooltip = res_tooltip,
      hover_options = list(fill_extrusion_color = "#ffffff")
    )

  # 3D: height=living_area, colour=year_built (hidden)
  m_res <- m_res |>
    add_fill_extrusion_layer(
      id = "res-3d-la",
      source = res_sf,
      visibility = "none",
      fill_extrusion_height = interpolate(
        column = "total_living_area",
        values = c(0, 500, 1000, 2000, 4000),
        stops  = c(0, 50, 150, 400, 800)
      ),
      fill_extrusion_color = interpolate(
        column = "year_built",
        values = yb_parcel_breaks,
        stops  = yb_parcel_colors
      ),
      fill_extrusion_opacity = 0.8,
      tooltip = res_tooltip,
      hover_options = list(fill_extrusion_color = "#ffffff")
    )
} else {
  m_res <- m_res |>
    add_pmtiles_source(
      id = "res-parcels-source",
      url = pmtiles_url
    )

  # 3D: height=year_built, colour=living_area (DEFAULT - visible)
  m_res <- m_res |>
    add_fill_extrusion_layer(
      id = "res-3d-yb",
      source = "res-parcels-source",
      source_layer = layer_name,
      filter = res_pmtiles_filter,
      fill_extrusion_height = interpolate(
        column = "year_built",
        values = c(1900, 1940, 1960, 1980, 2000, 2025),
        stops  = c(20, 80, 200, 400, 700, 1000)
      ),
      fill_extrusion_color = interpolate(
        column = "total_living_area",
        values = la_parcel_breaks,
        stops  = la_parcel_colors
      ),
      fill_extrusion_opacity = 0.8,
      tooltip = res_tooltip,
      hover_options = list(fill_extrusion_color = "#ffffff")
    )

  # 3D: height=living_area, colour=year_built (hidden)
  m_res <- m_res |>
    add_fill_extrusion_layer(
      id = "res-3d-la",
      source = "res-parcels-source",
      source_layer = layer_name,
      filter = res_pmtiles_filter,
      visibility = "none",
      fill_extrusion_height = interpolate(
        column = "total_living_area",
        values = c(0, 500, 1000, 2000, 4000),
        stops  = c(0, 50, 150, 400, 800)
      ),
      fill_extrusion_color = interpolate(
        column = "year_built",
        values = yb_parcel_breaks,
        stops  = yb_parcel_colors
      ),
      fill_extrusion_opacity = 0.8,
      tooltip = res_tooltip,
      hover_options = list(fill_extrusion_color = "#ffffff")
    )
}

# --- Legends ---

# Year built colour legend (used when colour = year built)
yb_legend_labels <- as.character(yb_parcel_breaks)
yb_legend_rows <- paste0(sapply(seq_along(yb_parcel_colors), function(i) {
  label <- if (i == 1) {
    paste0("&lt; ", yb_legend_labels[i + 1])
  } else if (i == length(yb_parcel_colors)) {
    paste0("&ge; ", yb_legend_labels[i])
  } else {
    paste0(yb_legend_labels[i], " \u2013 ", yb_legend_labels[i + 1])
  }
  paste0(
    '<div style="display:flex;align-items:center;margin-bottom:3px;">',
    '<span style="display:inline-block;width:18px;height:12px;background:', yb_parcel_colors[i],
    ';border-radius:2px;margin-right:6px;flex-shrink:0;"></span>',
    '<span>', label, '</span></div>'
  )
}), collapse = "")

# Living area colour legend (used when colour = living area)
la_legend_labels <- format(la_parcel_breaks, big.mark = ",")
la_legend_rows <- paste0(sapply(seq_along(la_parcel_colors), function(i) {
  label <- if (i == 1) {
    paste0("&lt; ", la_legend_labels[i + 1], " sq ft")
  } else if (i == length(la_parcel_colors)) {
    paste0("&ge; ", la_legend_labels[i], " sq ft")
  } else {
    paste0(la_legend_labels[i], " \u2013 ", la_legend_labels[i + 1], " sq ft")
  }
  paste0(
    '<div style="display:flex;align-items:center;margin-bottom:3px;">',
    '<span style="display:inline-block;width:18px;height:12px;background:', la_parcel_colors[i],
    ';border-radius:2px;margin-right:6px;flex-shrink:0;"></span>',
    '<span>', label, '</span></div>'
  )
}), collapse = "")

# Neighbourhood year built legend (quantile-based)
nb_yb_legend_labels <- as.character(round(yb_breaks))
nb_yb_legend_rows <- paste0(sapply(seq_along(yb_colors), function(i) {
  label <- if (i == 1) {
    paste0("&lt; ", nb_yb_legend_labels[i + 1])
  } else if (i == length(yb_colors)) {
    paste0("&ge; ", nb_yb_legend_labels[i])
  } else {
    paste0(nb_yb_legend_labels[i], " \u2013 ", nb_yb_legend_labels[i + 1])
  }
  paste0(
    '<div style="display:flex;align-items:center;margin-bottom:3px;">',
    '<span style="display:inline-block;width:18px;height:12px;background:', yb_colors[i],
    ';border-radius:2px;margin-right:6px;flex-shrink:0;"></span>',
    '<span>', label, '</span></div>'
  )
}), collapse = "")

# Neighbourhood living area legend (quantile-based)
nb_la_legend_labels <- format(round(la_breaks), big.mark = ",")
nb_la_legend_rows <- paste0(sapply(seq_along(la_colors), function(i) {
  label <- if (i == 1) {
    paste0("&lt; ", nb_la_legend_labels[i + 1], " sq ft")
  } else if (i == length(la_colors)) {
    paste0("&ge; ", nb_la_legend_labels[i], " sq ft")
  } else {
    paste0(nb_la_legend_labels[i], " \u2013 ", nb_la_legend_labels[i + 1], " sq ft")
  }
  paste0(
    '<div style="display:flex;align-items:center;margin-bottom:3px;">',
    '<span style="display:inline-block;width:18px;height:12px;background:', la_colors[i],
    ';border-radius:2px;margin-right:6px;flex-shrink:0;"></span>',
    '<span>', label, '</span></div>'
  )
}), collapse = "")

# Combined control with two radio groups
res_control_html <- paste0(
  '<div style="background:rgba(255,255,255,0.85);padding:10px 12px;border-radius:6px;color:#333;font-family:sans-serif;font-size:13px;line-height:1.6;border:1px solid #ccc;">',

  '<div style="font-weight:bold;margin-bottom:4px;">Metric</div>',
  '<label style="display:block;cursor:pointer;">',
  '<input type="radio" name="res-metric" value="year-built" checked style="margin-right:5px;">Year Built',
  '</label>',
  '<label style="display:block;cursor:pointer;">',
  '<input type="radio" name="res-metric" value="living-area" style="margin-right:5px;">Living Area',
  '</label>',

  '<div style="font-weight:bold;margin-top:8px;margin-bottom:4px;">View</div>',
  '<label style="display:block;cursor:pointer;">',
  '<input type="radio" name="res-view" value="3d" checked style="margin-right:5px;">3D Parcels',
  '</label>',
  '<label style="display:block;cursor:pointer;">',
  '<input type="radio" name="res-view" value="neighbourhood" style="margin-right:5px;">Neighbourhood',
  '</label>',

  # Legend: 3D year-built (height=year, colour=living area) -- DEFAULT visible
  '<div id="res-leg-3d-yb" style="margin-top:8px;padding-top:8px;border-top:1px solid rgba(0,0,0,0.2);font-size:12px;line-height:1.5;">',
  '<div style="font-weight:bold;margin-bottom:3px;">Height: Year Built</div>',
  '<div style="margin-bottom:5px;font-size:11px;color:#666;">Taller = newer</div>',
  '<div style="font-weight:bold;margin-bottom:3px;">Colour: Living Area</div>',
  la_legend_rows,
  '</div>',

  # Legend: 3D living-area (height=area, colour=year built)
  '<div id="res-leg-3d-la" style="display:none;margin-top:8px;padding-top:8px;border-top:1px solid rgba(0,0,0,0.2);font-size:12px;line-height:1.5;">',
  '<div style="font-weight:bold;margin-bottom:3px;">Height: Living Area</div>',
  '<div style="margin-bottom:5px;font-size:11px;color:#666;">Taller = larger</div>',
  '<div style="font-weight:bold;margin-bottom:3px;">Colour: Year Built</div>',
  yb_legend_rows,
  '</div>',

  # Legend: Neighbourhood year built
  '<div id="res-leg-nb-yb" style="display:none;margin-top:8px;padding-top:8px;border-top:1px solid rgba(0,0,0,0.2);font-size:12px;line-height:1.5;">',
  '<div style="font-weight:bold;margin-bottom:5px;">Median Year Built</div>',
  nb_yb_legend_rows,
  '</div>',

  # Legend: Neighbourhood living area
  '<div id="res-leg-nb-la" style="display:none;margin-top:8px;padding-top:8px;border-top:1px solid rgba(0,0,0,0.2);font-size:12px;line-height:1.5;">',
  '<div style="font-weight:bold;margin-bottom:5px;">Median Living Area</div>',
  nb_la_legend_rows,
  '</div>',

  '</div>'
)

m_res <- m_res |>
  add_control(res_control_html, position = "bottom-left") |>
  htmlwidgets::onRender("
    function(el, x) {
      var poll = setInterval(function() {
        if (!el.map || !el.map.isStyleLoaded()) return;
        clearInterval(poll);
        var map = el.map;

        var layerMap = {
          '3d-year-built':            'res-3d-yb',
          '3d-living-area':           'res-3d-la',
          'neighbourhood-year-built': 'res-nb-yearbuilt',
          'neighbourhood-living-area':'res-nb-livingarea'
        };
        var legendMap = {
          '3d-year-built':            document.getElementById('res-leg-3d-yb'),
          '3d-living-area':           document.getElementById('res-leg-3d-la'),
          'neighbourhood-year-built': document.getElementById('res-leg-nb-yb'),
          'neighbourhood-living-area':document.getElementById('res-leg-nb-la')
        };

        var allLayerIds = Object.values(layerMap);
        var allLegends  = Object.values(legendMap);

        function updateMap() {
          var metric = el.querySelector('input[name=\"res-metric\"]:checked').value;
          var view   = el.querySelector('input[name=\"res-view\"]:checked').value;
          var key    = view + '-' + metric;

          allLayerIds.forEach(function(id) {
            map.setLayoutProperty(id, 'visibility', 'none');
          });
          allLegends.forEach(function(lg) {
            if (lg) lg.style.display = 'none';
          });

          var activeLayer = layerMap[key];
          var activeLegend = legendMap[key];
          if (activeLayer) map.setLayoutProperty(activeLayer, 'visibility', 'visible');
          if (activeLegend) activeLegend.style.display = 'block';
        }

        el.querySelectorAll('input[name=\"res-metric\"]').forEach(function(r) {
          r.addEventListener('change', updateMap);
        });
        el.querySelectorAll('input[name=\"res-view\"]').forEach(function(r) {
          r.addEventListener('change', updateMap);
        });
      }, 200);
    }
  ")

m_res
```

### Residential Summary by Neighbourhood

```{r}
#| label: res-summary-table

res_summary <- res_sf |>
  st_drop_geometry() |>
  group_by(neighbourhood_area) |>
  summarise(
    parcels            = n(),
    median_year_built  = round(median(year_built, na.rm = TRUE)),
    median_living_area = round(median(total_living_area, na.rm = TRUE)),
    total_assessed     = sum(total_assessed_value, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(total_assessed))

reactable(
  res_summary,
  sortable = TRUE,
  defaultSorted = list(total_assessed = "desc"),
  defaultPageSize = 25,
  showPageSizeOptions = TRUE,
  pageSizeOptions = c(25, 50, 100),
  striped = TRUE,
  highlight = TRUE,
  compact = TRUE,
  searchable = TRUE,
  columns = list(
    neighbourhood_area = colDef(name = "Neighbourhood", minWidth = 170),
    parcels = colDef(
      name = "Parcels",
      format = colFormat(separators = TRUE),
      defaultSortOrder = "desc"
    ),
    median_year_built = colDef(
      name = "Median Year Built",
      format = colFormat(digits = 0),
      defaultSortOrder = "desc"
    ),
    median_living_area = colDef(
      name = "Median Living Area (sq ft)",
      format = colFormat(separators = TRUE, digits = 0),
      defaultSortOrder = "desc"
    ),
    total_assessed = colDef(
      name = "Total Assessment",
      format = colFormat(prefix = "$", separators = TRUE, digits = 0),
      defaultSortOrder = "desc",
      minWidth = 145
    )
  )
)
```

