---
title: "Winnipeg Assessment Parcels — 3D Value Map"
author: "Jay Schellenberg"
date: today
date-format: "MMMM D, YYYY"
format:
  html:
    toc: true
output-file: index.html
execute:
  echo: false
  warning: false
  message: false
---

## Overview

Interactive 3D map of Winnipeg's property assessment parcels.
**Height** encodes total assessed value; **color** encodes neighbourhood.
Use the **search bar** (top left) to find any address in Winnipeg.
Hover over a parcel to see its assessed value, roll number, and how it compares to the neighbourhood median.
Use the **layer toggle** (bottom left) to switch to a neighbourhood-level view comparing total assessed value across neighbourhoods.

```{r}
#| label: setup
#| include: false

library(duckdb)
library(DBI)
library(freestiler)
library(mapgl)
library(arrow)
library(geoarrow)
library(dplyr)
library(sf)
library(ggplot2)
library(ggspatial)
library(scico)
library(gt)
library(fs)
library(glue)
library(cli)
library(scales)
library(readr)
library(stringr)

fs::dir_create("exports")

layer_name <- "parcels"
pmtiles_file <- "data/assessment-parcels.pmtiles"

# Set to TRUE to use local data embedded in the HTML (works offline, larger file).
# Set to FALSE to use PMTiles from docs/ (smaller file, requires Git LFS).
use_local_data <- FALSE

# PMTiles URL for the deployed map — relative path served from docs/ alongside index.html
pmtiles_url <- "assessment-parcels.pmtiles"

# Assessment roll metadata
base_value_date <- "April 1, 2023"
```

```{r}
#| label: find-data
#| include: false

parquet_files <- fs::dir_ls("data", glob = "*/assessment-parcels-*.parquet")

if (length(parquet_files) == 0) {
  stop("No GeoParquet files found in data/. Run WpgAssessmentParcels.qmd first.")
}

input_file <- sort(parquet_files, decreasing = TRUE)[1]
file_date <- stringr::str_extract(basename(input_file), "\\d{4}-\\d{2}-\\d{2}")

# Extract assessment year from the current_assessment_year field
asmt_year <- read_parquet(input_file, col_select = "current_assessment_year") |>
  pull(current_assessment_year) |>
  na.omit() |>
  head(1) |>
  as.character()

cli_alert_success("Using: {basename(input_file)}")
cli_alert_info("Download date: {file_date}")
cli_alert_info("File size: {fs::file_size(input_file)}")
```

## City Overview

```{r}
#| label: city-overview
#| include: false

parcels_all <- read_parquet(input_file) |> as.data.frame()

n_parcels <- nrow(parcels_all)
n_neighbourhoods <- n_distinct(parcels_all$neighbourhood_area)
total_assessed <- sum(parcels_all$total_assessed_value, na.rm = TRUE)
median_assessed <- median(parcels_all$total_assessed_value, na.rm = TRUE)

# Top neighbourhood by total assessed value
top_neighbourhood <- parcels_all |>
  group_by(neighbourhood_area) |>
  summarise(total = sum(total_assessed_value, na.rm = TRUE), .groups = "drop") |>
  slice_max(total, n = 1)
```

Winnipeg's assessment roll (`r asmt_year` — `r base_value_date` reference date) covers **`r format(n_parcels, big.mark = ",")`** property parcels across **`r n_neighbourhoods`** neighbourhoods, with a combined assessed value of **`r dollar(total_assessed, accuracy = 1, scale = 1e-9, suffix = "B")`**. The median property assessment is **`r dollar(median_assessed, accuracy = 1)`**. The largest neighbourhood by total assessed value is **`r top_neighbourhood$neighbourhood_area`** at `r dollar(top_neighbourhood$total, accuracy = 1, scale = 1e-9, suffix = "B")`.

Data source: [City of Winnipeg Assessment Parcels](https://data.winnipeg.ca/Assessment-Taxation-Corporate/Assessment-Parcels/d4mq-wa44) (downloaded `r file_date`).

```{r}
#| label: load-map-data
#| include: false

cli_alert_info("Loading parcel geometry from {basename(input_file)}...")

parcels_sf <- read_parquet(input_file) |>
  mutate(geometry = st_as_sfc(geometry)) |>
  st_as_sf()

# Simplify geometry for faster rendering and smaller HTML output
# dTolerance = 0.0002 (~22m) keeps parcels recognizable at street zoom
if (use_local_data) {
  sf_use_s2(FALSE)
  parcels_sf <- st_make_valid(parcels_sf) |>
    st_simplify(dTolerance = 0.0002, preserveTopology = TRUE)
  sf_use_s2(TRUE)
  cli_alert_info("Simplified geometry for local embedding")
}

# Compute neighbourhood medians and join back for tooltip context
neighbourhood_medians <- parcels_sf |>
  st_drop_geometry() |>
  group_by(neighbourhood_area) |>
  summarise(Neighbourhood_Median = median(total_assessed_value, na.rm = TRUE), .groups = "drop")

parcels_sf <- parcels_sf |>
  left_join(neighbourhood_medians, by = "neighbourhood_area") |>
  mutate(
    Neighbourhood_Median_Fmt = if_else(
      is.na(Neighbourhood_Median), "N/A", dollar(Neighbourhood_Median, accuracy = 1)
    ),
    Total_Value_Fmt = dollar(total_assessed_value, accuracy = 1)
  ) |>
  select(roll_number, total_assessed_value, Total_Value_Fmt,
         Neighbourhood_Median, Neighbourhood_Median_Fmt,
         full_address, neighbourhood_area,
         property_use_code, building_type, year_built,
         total_living_area, dwelling_units)

cli_alert_success("Loaded {format(nrow(parcels_sf), big.mark = ',')} parcels")

# Dissolve parcels into neighbourhood boundaries for choropleth layer
cli_alert_info("Dissolving parcels by neighbourhood...")
sf_use_s2(FALSE)
parcels_sf <- st_make_valid(parcels_sf)
neighbourhood_sf <- parcels_sf |>
  group_by(neighbourhood_area) |>
  summarise(
    Total_Assessed = sum(total_assessed_value, na.rm = TRUE),
    Parcel_Count = n(),
    Median_Value = median(total_assessed_value, na.rm = TRUE),
    .groups = "drop"
  ) |>
  st_make_valid() |>
  st_simplify(dTolerance = 0.001, preserveTopology = TRUE) |>
  st_make_valid() |>
  mutate(
    Total_Assessed_Fmt = dollar(Total_Assessed, accuracy = 1),
    Median_Value_Fmt = dollar(Median_Value, accuracy = 1),
    Value_Rank = paste(rank(-Total_Assessed, ties.method = "min"), "of", n())
  )
sf_use_s2(TRUE)
cli_alert_success("Created {nrow(neighbourhood_sf)} neighbourhood polygons")
```

```{r}
#| label: generate-pmtiles
#| include: false
#| eval: !expr '!use_local_data'

# Skip if PMTiles is newer than source parquet
needs_rebuild <- !fs::file_exists(pmtiles_file) ||
  fs::file_info(pmtiles_file)$modification_time <
  fs::file_info(input_file)$modification_time

if (needs_rebuild) {
  cli_alert_info("Generating PMTiles from {basename(input_file)}...")

  freestile(
    input = parcels_sf,
    output = pmtiles_file,
    layer_name = layer_name,
    min_zoom = 8L,
    max_zoom = 14L
  )

  cli_alert_success("PMTiles written: {fs::file_size(pmtiles_file)}")
} else {
  cli_alert_info("PMTiles up to date: {fs::file_size(pmtiles_file)}")
}
```

```{r}
#| label: neighbourhood-codes
#| include: false

parcels_df <- parcels_all |>
  count(neighbourhood_area, sort = TRUE)

cli_alert_info("Found {nrow(parcels_df)} neighbourhoods")
```

```{r}
#| label: define-palette
#| include: false

# Define a categorical color palette for the top neighbourhoods
# Colors chosen for visual distinction on a dark basemap
palette <- c(
  "#e74c3c",   # red
  "#3498db",   # blue
  "#2ecc71",   # green
  "#f39c12",   # orange
  "#9b59b6",   # purple
  "#1abc9c",   # teal
  "#e67e22",   # dark orange
  "#ec407a",   # pink
  "#26c6da",   # cyan
  "#ffca28",   # amber
  "#8d6e63",   # brown
  "#78909c"    # blue-grey
)

# Use top neighbourhoods by parcel count for color coding
neighbourhood_names <- parcels_df$neighbourhood_area
neighbourhood_colors <- rep_len(palette, length(neighbourhood_names))
```

## 3D Map

```{r}
#| label: map-3d

# Quantile breaks for neighbourhood choropleth color scale
total_vals <- neighbourhood_sf$Total_Assessed[!is.na(neighbourhood_sf$Total_Assessed)]
q_breaks <- as.numeric(quantile(total_vals, probs = c(0, 0.2, 0.4, 0.6, 0.8, 1.0)))

m <- maplibre(
  style = carto_style("dark-matter"),
  center = c(-97.14, 49.89),
  zoom = 11,
  pitch = 30,
  bearing = 0,
  projection = "mercator"
) |>
  add_navigation_control(position = "top-right") |>
  add_geocoder_control(
    position = "top-left",
    placeholder = "Search for an address..."
  )

# Neighbourhood choropleth layer (hidden by default)
m <- m |>
  add_fill_layer(
    id = "neighbourhood-total-value",
    source = neighbourhood_sf,
    fill_color = interpolate(
      column = "Total_Assessed",
      values = q_breaks,
      stops = c("#2166ac", "#67a9cf", "#d1e5f0", "#fddbc7", "#ef8a62", "#b2182b")
    ),
    fill_opacity = 0.8,
    visibility = "none",
    tooltip = concat(
      "<strong>", get_column("neighbourhood_area"), "</strong><br>",
      "Total Assessed Value: ", get_column("Total_Assessed_Fmt"), "<br>",
      "Rank: ", get_column("Value_Rank"), "<br>",
      "Parcels: ", get_column("Parcel_Count"), "<br>",
      "Median Value: ", get_column("Median_Value_Fmt")
    ),
    hover_options = list(
      fill_color = "#ffffff",
      fill_opacity = 1
    )
  )

# 3D parcel extrusion layer (visible by default)
if (use_local_data) {
  m <- m |>
    add_fill_extrusion_layer(
      id = "wpg-parcels-3d",
      source = parcels_sf,
      fill_extrusion_height = interpolate(
        column = "total_assessed_value",
        values = c(0, 100000, 500000, 2000000, 10000000),
        stops = c(0, 50, 200, 600, 2000)
      ),
      fill_extrusion_color = match_expr(
        column = "neighbourhood_area",
        values = neighbourhood_names,
        stops = neighbourhood_colors,
        default = "#95a5a6"
      ),
      fill_extrusion_opacity = 0.8,
      tooltip = concat(
        "<strong>", get_column("full_address"), "</strong><br>",
        "Assessed Value: ", get_column("Total_Value_Fmt"), "<br>",
        "Neighbourhood Median: ", get_column("Neighbourhood_Median_Fmt"), "<br>",
        "Neighbourhood: ", get_column("neighbourhood_area"), "<br>",
        "Roll No: ", get_column("roll_number"), "<br>",
        "Use: ", get_column("property_use_code"), "<br>",
        "Dwelling Units: ", get_column("dwelling_units")
      ),
      hover_options = list(
        fill_extrusion_color = "#ffffff"
      )
    )
} else {
  m <- m |>
    add_pmtiles_source(
      id = "wpg-parcels-source",
      url = pmtiles_url
    ) |>
    add_fill_extrusion_layer(
      id = "wpg-parcels-3d",
      source = "wpg-parcels-source",
      source_layer = layer_name,
      fill_extrusion_height = interpolate(
        column = "total_assessed_value",
        values = c(0, 100000, 500000, 2000000, 10000000),
        stops = c(0, 50, 200, 600, 2000)
      ),
      fill_extrusion_color = match_expr(
        column = "neighbourhood_area",
        values = neighbourhood_names,
        stops = neighbourhood_colors,
        default = "#95a5a6"
      ),
      fill_extrusion_opacity = 0.8,
      tooltip = concat(
        "<strong>", get_column("full_address"), "</strong><br>",
        "Assessed Value: ", get_column("Total_Value_Fmt"), "<br>",
        "Neighbourhood Median: ", get_column("Neighbourhood_Median_Fmt"), "<br>",
        "Neighbourhood: ", get_column("neighbourhood_area"), "<br>",
        "Roll No: ", get_column("roll_number"), "<br>",
        "Use: ", get_column("property_use_code"), "<br>",
        "Dwelling Units: ", get_column("dwelling_units")
      ),
      hover_options = list(
        fill_extrusion_color = "#ffffff"
      )
    )
}

# Build legend with abbreviated breakpoint labels (e.g. $1.2B, $450M)
legend_colors <- c("#2166ac", "#67a9cf", "#d1e5f0", "#fddbc7", "#ef8a62", "#b2182b")
fmt_short <- function(x) {
  dplyr::case_when(
    x >= 1e9  ~ paste0("$", trimws(format(round(x / 1e9, 1), nsmall = 1)), "B"),
    x >= 1e6  ~ paste0("$", trimws(format(round(x / 1e6, 0))), "M"),
    x >= 1e3  ~ paste0("$", trimws(format(round(x / 1e3, 0))), "K"),
    TRUE      ~ paste0("$", trimws(format(round(x, 0))))
  )
}
legend_labels <- fmt_short(q_breaks)

legend_rows <- paste0(sapply(seq_along(legend_colors), function(i) {
  label <- if (i == 1) {
    paste0("&lt; ", legend_labels[i + 1])
  } else if (i == length(legend_colors)) {
    paste0("&ge; ", legend_labels[i])
  } else {
    paste0(legend_labels[i], " \u2013 ", legend_labels[i + 1])
  }
  paste0(
    '<div style="display:flex;align-items:center;margin-bottom:3px;">',
    '<span style="display:inline-block;width:18px;height:12px;background:', legend_colors[i],
    ';border-radius:2px;margin-right:6px;flex-shrink:0;"></span>',
    '<span>', label, '</span></div>'
  )
}), collapse = "")

# Combine layer toggle + legend into ONE add_control (mapgl only keeps one custom control)
combined_control_html <- paste0(
  '<div style="background:rgba(0,0,0,0.75);padding:10px 12px;border-radius:6px;color:#fff;font-family:sans-serif;font-size:13px;line-height:1.6;">',
  '<div style="font-weight:bold;margin-bottom:4px;">View</div>',
  '<label style="display:block;cursor:pointer;">',
  '<input type="radio" name="map-layer" value="3d" checked style="margin-right:5px;">3D Parcels',
  '</label>',
  '<label style="display:block;cursor:pointer;">',
  '<input type="radio" name="map-layer" value="neighbourhood" style="margin-right:5px;">Total Value by Neighbourhood',
  '</label>',
  '<div id="neighbourhood-legend" style="display:none;margin-top:8px;padding-top:8px;border-top:1px solid rgba(255,255,255,0.3);font-size:12px;line-height:1.5;">',
  '<div style="font-weight:bold;margin-bottom:5px;">Total Assessed Value</div>',
  legend_rows,
  '</div>',
  '</div>'
)

m <- m |>
  add_control(combined_control_html, position = "bottom-left") |>
  htmlwidgets::onRender("
    function(el, x) {
      var poll = setInterval(function() {
        if (!el.map || !el.map.isStyleLoaded()) return;
        clearInterval(poll);
        var map = el.map;
        var legend = document.getElementById('neighbourhood-legend');
        var radios = el.querySelectorAll('input[name=\"map-layer\"]');
        radios.forEach(function(radio) {
          radio.addEventListener('change', function() {
            if (this.value === '3d') {
              map.setLayoutProperty('wpg-parcels-3d', 'visibility', 'visible');
              map.setLayoutProperty('neighbourhood-total-value', 'visibility', 'none');
              if (legend) legend.style.display = 'none';
            } else {
              map.setLayoutProperty('wpg-parcels-3d', 'visibility', 'none');
              map.setLayoutProperty('neighbourhood-total-value', 'visibility', 'visible');
              if (legend) legend.style.display = 'block';
            }
          });
        });
      }, 200);
    }
  ")

m
```

## Assessed Value by Neighbourhood (`r asmt_year`)

```{r}
#| label: summary-stats

summary_table <- parcels_all |>
  group_by(neighbourhood_area) |>
  summarise(
    count = n(),
    total_value = sum(total_assessed_value, na.rm = TRUE),
    median_value = median(total_assessed_value, na.rm = TRUE),
    mean_value = mean(total_assessed_value, na.rm = TRUE),
    max_value = if (all(is.na(total_assessed_value))) NA_real_
                else max(total_assessed_value, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(total_value))

summary_display <- summary_table |>
  mutate(
    total_value = dollar(total_value, accuracy = 1),
    median_value = dollar(median_value, accuracy = 1),
    mean_value = dollar(mean_value, accuracy = 1),
    max_value = dollar(max_value, accuracy = 1)
  )

summary_display |> knitr::kable(
  col.names = c("Neighbourhood", "Parcels", "Total Assessed Value",
                "Median", "Mean", "Max"),
  align = c("l", "r", "r", "r", "r", "r")
)
```

```{r}
#| label: compute-change
#| include: false

# Compute per-parcel % change from current (2026) to proposed (2027)
change_stats <- parcels_all |>
  filter(
    !is.na(total_assessed_value),
    !is.na(total_proposed_assessment_value),
    total_assessed_value > 0
  ) |>
  mutate(
    pct_change = (total_proposed_assessment_value - total_assessed_value) /
                  total_assessed_value * 100
  )

n_change_parcels <- nrow(change_stats)
citywide_median_change <- median(change_stats$pct_change, na.rm = TRUE)

# Summarize by neighbourhood
change_by_neighbourhood <- change_stats |>
  group_by(neighbourhood_area) |>
  summarise(
    change_parcel_count = n(),
    median_pct          = median(pct_change, na.rm = TRUE),
    mean_pct            = mean(pct_change, na.rm = TRUE),
    iqr_pct             = IQR(pct_change, na.rm = TRUE),
    q25_pct             = quantile(pct_change, 0.25, na.rm = TRUE),
    q75_pct             = quantile(pct_change, 0.75, na.rm = TRUE),
    .groups = "drop"
  )

# Join change stats to existing dissolved neighbourhood boundaries
change_neighbourhood_sf <- neighbourhood_sf |>
  left_join(change_by_neighbourhood, by = "neighbourhood_area") |>
  mutate(
    median_pct_fmt = paste0(round(median_pct, 1), "%"),
    mean_pct_fmt   = paste0(round(mean_pct, 1), "%"),
    iqr_pct_fmt    = paste0(round(iqr_pct, 1), "%")
  )

cli_alert_info("Citywide median change: {round(citywide_median_change, 1)}%")
```

## Assessment Change — 2026 → 2027 (Proposed)

This section compares Winnipeg's **current (`r asmt_year`)** property assessments to the **proposed (2027)** values across `r format(n_change_parcels, big.mark = ",")` parcels. The citywide median assessment change is **`r round(citywide_median_change, 1)`%**.

### Where Assessments Jumped Most

Median percentage change in assessed value by neighbourhood. The diverging scale is centred on the citywide median (`r round(citywide_median_change, 1)`%).

#### Interactive Map

```{r}
#| label: map-a-mapgl

# Diverging breaks centred on citywide median
median_vals <- change_neighbourhood_sf$median_pct[!is.na(change_neighbourhood_sf$median_pct)]
val_range <- range(median_vals)
max_dev <- max(abs(val_range - citywide_median_change))
break_low  <- citywide_median_change - max_dev
break_high <- citywide_median_change + max_dev
q_breaks_a <- c(break_low,
                citywide_median_change - max_dev * 0.5,
                citywide_median_change,
                citywide_median_change + max_dev * 0.5,
                break_high)

maplibre(
  style = carto_style("dark-matter"),
  center = c(-97.14, 49.89),
  zoom = 11,
  pitch = 0,
  bearing = 0,
  projection = "mercator"
) |>
  add_navigation_control(position = "top-right") |>
  add_fill_layer(
    id = "median-change",
    source = change_neighbourhood_sf,
    fill_color = interpolate(
      column = "median_pct",
      values = q_breaks_a,
      stops = c("#2166ac", "#92c5de", "#f7f7f7", "#f4a582", "#b2182b")
    ),
    fill_opacity = 0.85,
    tooltip = concat(
      "<strong>", get_column("neighbourhood_area"), "</strong><br>",
      "Median Change: ", get_column("median_pct_fmt"), "<br>",
      "Mean Change: ", get_column("mean_pct_fmt"), "<br>",
      "Parcels: ", get_column("change_parcel_count")
    ),
    hover_options = list(
      fill_color = "#ffffff",
      fill_opacity = 1
    )
  ) |>
  add_line_layer(
    id = "neighbourhood-borders-a",
    source = change_neighbourhood_sf,
    line_color = "#ffffff",
    line_width = 0.5,
    line_opacity = 0.3
  )
```

#### Static Map (for export)

```{r}
#| label: map-a-ggplot
#| fig-width: 10
#| fig-height: 10
#| dpi: 150

p_a <- ggplot(change_neighbourhood_sf) +
  geom_sf(aes(fill = median_pct), color = "grey30", linewidth = 0.15) +
  scico::scale_fill_scico(
    palette = "vik",
    midpoint = citywide_median_change,
    name = "Median % Change",
    labels = function(x) paste0(round(x, 0), "%")
  ) +
  annotation_scale(
    location = "bl",
    width_hint = 0.25,
    style = "ticks",
    text_col = "grey30",
    line_col = "grey30"
  ) +
  annotation_north_arrow(
    location = "tr",
    which_north = "true",
    height = unit(1, "cm"),
    width = unit(1, "cm"),
    style = north_arrow_minimal(text_col = "grey30", line_col = "grey30")
  ) +
  labs(
    title = "Where Assessments Jumped Most",
    subtitle = glue("Median % change in assessed value by neighbourhood (2026 \u2192 2027 proposed)\nCitywide median: {round(citywide_median_change, 1)}%"),
    caption = glue("Source: City of Winnipeg Open Data \u2022 Downloaded {file_date}")
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 11, color = "grey40"),
    plot.caption = element_text(size = 8, color = "grey50"),
    legend.position = "bottom",
    legend.key.width = unit(2.5, "cm"),
    legend.key.height = unit(0.4, "cm")
  )

p_a

ggsave("exports/assessment-change-median.png", p_a,
       width = 10, height = 10, dpi = 150, bg = "white")
```

### Assessment Volatility

Interquartile range (IQR) of percentage change within each neighbourhood. Higher IQR indicates greater spread — some parcels saw large increases while others stayed flat.

#### Interactive Map

```{r}
#| label: map-b-mapgl

iqr_vals <- change_neighbourhood_sf$iqr_pct[!is.na(change_neighbourhood_sf$iqr_pct)]
q_breaks_b <- unique(as.numeric(quantile(iqr_vals, probs = c(0, 0.25, 0.5, 0.75, 1.0))))
if (length(q_breaks_b) < 2) {
  q_breaks_b <- c(min(iqr_vals), max(iqr_vals))
}
iqr_colors <- c("#fef0d9", "#fdcc8a", "#fc8d59", "#d7301f")[seq_len(length(q_breaks_b))]

maplibre(
  style = carto_style("dark-matter"),
  center = c(-97.14, 49.89),
  zoom = 11,
  pitch = 0,
  bearing = 0,
  projection = "mercator"
) |>
  add_navigation_control(position = "top-right") |>
  add_fill_layer(
    id = "iqr-change",
    source = change_neighbourhood_sf,
    fill_color = interpolate(
      column = "iqr_pct",
      values = q_breaks_b,
      stops = iqr_colors
    ),
    fill_opacity = 0.85,
    tooltip = concat(
      "<strong>", get_column("neighbourhood_area"), "</strong><br>",
      "IQR of Change: ", get_column("iqr_pct_fmt"), "<br>",
      "Q25: ", get_column("q25_pct"), "% \u2013 Q75: ", get_column("q75_pct"), "%<br>",
      "Median Change: ", get_column("median_pct_fmt"), "<br>",
      "Parcels: ", get_column("change_parcel_count")
    ),
    hover_options = list(
      fill_color = "#ffffff",
      fill_opacity = 1
    )
  ) |>
  add_line_layer(
    id = "neighbourhood-borders-b",
    source = change_neighbourhood_sf,
    line_color = "#ffffff",
    line_width = 0.5,
    line_opacity = 0.3
  )
```

#### Static Map (for export)

```{r}
#| label: map-b-ggplot
#| fig-width: 10
#| fig-height: 10
#| dpi: 150

p_b <- ggplot(change_neighbourhood_sf) +
  geom_sf(aes(fill = iqr_pct), color = "grey30", linewidth = 0.15) +
  scico::scale_fill_scico(
    palette = "lajolla",
    direction = 1,
    name = "IQR of % Change",
    labels = function(x) paste0(round(x, 0), "%")
  ) +
  annotation_scale(
    location = "bl",
    width_hint = 0.25,
    style = "ticks",
    text_col = "grey30",
    line_col = "grey30"
  ) +
  annotation_north_arrow(
    location = "tr",
    which_north = "true",
    height = unit(1, "cm"),
    width = unit(1, "cm"),
    style = north_arrow_minimal(text_col = "grey30", line_col = "grey30")
  ) +
  labs(
    title = "Assessment Volatility by Neighbourhood",
    subtitle = glue("IQR of % change in assessed value (2026 \u2192 2027 proposed)\nHigher IQR = greater spread within neighbourhood"),
    caption = glue("Source: City of Winnipeg Open Data \u2022 Downloaded {file_date}")
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 11, color = "grey40"),
    plot.caption = element_text(size = 8, color = "grey50"),
    legend.position = "bottom",
    legend.key.width = unit(2.5, "cm"),
    legend.key.height = unit(0.4, "cm")
  )

p_b

ggsave("exports/assessment-volatility-iqr.png", p_b,
       width = 10, height = 10, dpi = 150, bg = "white")
```

### Top & Bottom Neighbourhoods by Median Assessment Change

```{r}
#| label: gt-table

top_15 <- change_by_neighbourhood |>
  slice_max(median_pct, n = 15) |>
  mutate(rank_group = "Largest Increases")

bottom_15 <- change_by_neighbourhood |>
  slice_min(median_pct, n = 15) |>
  arrange(median_pct) |>
  mutate(rank_group = "Smallest Increases")

bind_rows(top_15, bottom_15) |>
  select(rank_group, neighbourhood_area, median_pct, mean_pct, iqr_pct, change_parcel_count) |>
  gt(groupname_col = "rank_group") |>
  tab_header(
    title = glue("Neighbourhood Assessment Changes \u2014 {asmt_year} \u2192 2027 (Proposed)"),
    subtitle = glue("Citywide median change: {round(citywide_median_change, 1)}%")
  ) |>
  cols_label(
    neighbourhood_area = "Neighbourhood",
    median_pct = "Median %",
    mean_pct = "Mean %",
    iqr_pct = "IQR %",
    change_parcel_count = "Parcels"
  ) |>
  fmt_number(
    columns = c(median_pct, mean_pct, iqr_pct),
    decimals = 1,
    pattern = "{x}%"
  ) |>
  fmt_integer(columns = change_parcel_count) |>
  data_color(
    columns = median_pct,
    fn = scales::col_numeric(
      palette = c("#2166ac", "#f7f7f7", "#b2182b"),
      domain = range(change_by_neighbourhood$median_pct, na.rm = TRUE)
    )
  ) |>
  tab_source_note(
    source_note = glue("Source: City of Winnipeg Open Data \u2022 Downloaded {file_date}")
  ) |>
  tab_options(
    heading.title.font.size = 16,
    heading.subtitle.font.size = 12,
    table.font.size = 11,
    row_group.font.weight = "bold"
  )
```
