---
title: "Download Winnipeg Assessment Parcels"
subtitle: "Socrata SODA API — Full Dataset with Geometry"
author: "WpgOpenData"
date: today
format:
  html:
    toc: true
    code-fold: show
    code-summary: "Show code"
    embed-resources: true
execute:
  eval: false
  echo: true
  warning: true
  message: true
---

## Overview

Downloads all ~245,000 assessment parcel records (with MultiPolygon geometry)
from the [City of Winnipeg Open Data portal](https://data.winnipeg.ca/Assessment-Taxation-Corporate/Assessment-Parcels/d4mq-wa44/about_data)
via the Socrata SODA API. Saves as a date-stamped GeoParquet file.

The API defaults to 1,000 rows per request. We use `$limit=50000` with
`$offset` pagination to download all rows in ~5 batches.

## Setup

```{r}
#| label: setup

library(httr2)
library(readr)
library(dplyr)
library(sf)
library(arrow)
library(geoarrow)
library(cli)
library(fs)
library(glue)
library(ggplot2)
```

## Configuration

```{r}
#| label: config

base_url <- "https://data.winnipeg.ca/resource/d4mq-wa44"
batch_size <- 50000
output_dir <- "data"
force_download <- FALSE

# Create output directory
fs::dir_create(output_dir)

# Output file path with today's date
output_file <- fs::path(output_dir, glue("assessment-parcels-{Sys.Date()}.parquet"))

# Check if today's file already exists
if (fs::file_exists(output_file) && !force_download) {
  cli_alert_warning("File already exists: {output_file}")
  cli_alert_info("Set force_download <- TRUE to re-download")
}
```

## Query Row Count

```{r}
#| label: row-count

count_resp <- request(glue("{base_url}.json")) |>
  req_url_query(`$select` = "count(*) as total") |>
  req_retry(max_tries = 3, backoff = ~10) |>
  req_perform()

total_rows <- count_resp |>
  resp_body_json() |>
  purrr::pluck(1, "total") |>
  as.integer()

n_batches <- ceiling(total_rows / batch_size)

cli_alert_info("Total rows: {format(total_rows, big.mark = ',')}")
cli_alert_info("Batch size: {format(batch_size, big.mark = ',')}")
cli_alert_info("Batches needed: {n_batches}")
```

## Download Batches

```{r}
#| label: download

download_batch <- function(offset, batch_num, n_batches) {
  cli_alert_info(
    "Downloading batch {batch_num}/{n_batches} (offset {format(offset, big.mark = ',')})"
  )

  resp <- request(glue("{base_url}.csv")) |>
    req_url_query(
      `$limit` = batch_size,
      `$offset` = offset,
      `$order` = ":id"
    ) |>
    req_retry(max_tries = 3, backoff = ~10) |>
    req_throttle(rate = 1 / 2) |>
    req_perform()

  raw_csv <- resp_body_string(resp)
  batch_df <- read_csv(raw_csv, show_col_types = FALSE,
                       col_types = cols(roll_number = col_character()))

  probs <- problems(batch_df)
  if (nrow(probs) > 0) {
    cli_alert_warning("Batch {batch_num}: {nrow(probs)} parsing issue(s)")
    print(probs)
  }

  cli_alert_success(
    "Batch {batch_num}: {format(nrow(batch_df), big.mark = ',')} rows, {ncol(batch_df)} columns"
  )

  batch_df
}

# Skip download if file exists and not forcing
if (!fs::file_exists(output_file) || force_download) {
  offsets <- seq(0, by = batch_size, length.out = n_batches)

  batches <- lapply(seq_along(offsets), function(i) {
    download_batch(offsets[i], i, n_batches)
  })

  cli_alert_success("All {n_batches} batches downloaded")
} else {
  cli_alert_info("Skipping download — file exists: {output_file}")
}
```

## Combine and Convert to Spatial

```{r}
#| label: combine

if (!fs::file_exists(output_file) || force_download) {
  combined_df <- bind_rows(batches)

  cli_alert_info("Combined: {format(nrow(combined_df), big.mark = ',')} rows")

  # Check geometry column
  n_missing_geom <- sum(is.na(combined_df$geometry) | combined_df$geometry == "")
  if (n_missing_geom > 0) {
    cli_alert_warning("{n_missing_geom} rows have missing geometry")
  }

  # Filter to rows with valid geometry for sf conversion
  has_geom <- !is.na(combined_df$geometry) & combined_df$geometry != ""
  df_with_geom <- combined_df[has_geom, ]
  df_no_geom <- combined_df[!has_geom, ]

  # Convert WKT geometry to sf
  cli_alert_info("Converting WKT geometry to sf object...")
  parcels_sf <- st_as_sf(df_with_geom, wkt = "geometry", crs = 4326)

  # Validate row count
  expected <- total_rows
  actual <- nrow(parcels_sf) + nrow(df_no_geom)
  if (abs(actual - expected) > 100) {
    cli_alert_warning(
      "Row count mismatch: expected ~{format(expected, big.mark = ',')} but got {format(actual, big.mark = ',')}"
    )
  } else {
    cli_alert_success(
      "Row count validated: {format(actual, big.mark = ',')} rows (expected ~{format(expected, big.mark = ',')})"
    )
  }

  cli_alert_success("sf object created with {format(nrow(parcels_sf), big.mark = ',')} features")
}
```

## Write GeoParquet

```{r}
#| label: write-parquet

if (!fs::file_exists(output_file) || force_download) {
  cli_alert_info("Writing GeoParquet: {output_file}")
  write_parquet(parcels_sf, output_file)

  file_size <- fs::file_size(output_file)
  cli_alert_success("GeoParquet written: {file_size}")

  # Save rows without geometry separately if any exist
  if (nrow(df_no_geom) > 0) {
    no_geom_file <- fs::path(output_dir, glue("assessment-parcels-no-geometry-{Sys.Date()}.csv"))
    write_csv(df_no_geom, no_geom_file)
    cli_alert_info("{nrow(df_no_geom)} rows without geometry saved to {no_geom_file}")
  }
}
```

## Summary

```{r}
#| label: summary

# Read back the parquet to verify
# geoarrow_vctr geometry must be converted to sfc before sf functions work
parcels_sf <- read_parquet(output_file) |>
  mutate(geometry = st_as_sfc(geometry)) |>
  st_as_sf()

cli_h1("Dataset Summary")
cli_alert_info("Rows: {format(nrow(parcels_sf), big.mark = ',')}")
cli_alert_info("Columns: {ncol(parcels_sf)}")
cli_alert_info("CRS: {st_crs(parcels_sf)$input}")
cli_alert_info("Bounding box:")
print(st_bbox(parcels_sf))
cli_alert_info("File size: {fs::file_size(output_file)}")

glimpse(parcels_sf)
```

## Quick Validation Map

A random sample of 1,000 parcels to visually confirm geometry is correct.

```{r}
#| label: validation-map
#| fig-width: 10
#| fig-height: 8

set.seed(42)
sample_parcels <- parcels_sf |> slice_sample(n = 1000)

ggplot(sample_parcels) +
  geom_sf(aes(fill = total_assessed_value), color = NA, alpha = 0.7) +
  scale_fill_viridis_c(
    name = "Total Assessed Value",
    labels = scales::dollar,
    trans = "log10"
  ) +
  labs(
    title = "Winnipeg Assessment Parcels (Random Sample, n=1,000)",
    subtitle = glue("Downloaded {Sys.Date()} — {format(nrow(parcels_sf), big.mark = ',')} total parcels"),
    caption = "Source: City of Winnipeg Open Data"
  ) +
  theme_minimal()
```
